Title

### 2015-07-26 main.tex
minor edits
a few more tweaks

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many biological 'omics` applications require us to store, access, and analyze
large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have teamed up to bring the GATK (Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research and enterprise (e.g., DNANexus)~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers, and provides cost-saving benefits when computing resources are
not needed around the clock, it is no panacea.
First and foremost, cloud computing does not truly address the problem posed by
a corpus of data that is growing faster than Moore's law, because the computer
systems that make up those cloud datacenters are still themselves bound by
improvements in semiconductor technology; they still follow the trajectory
suggested by Moore's law.
Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.

@@

In the central dogma of molecular biology, DNA is transcribed into RNA, which
is translated by the ribosome into a polypeptide chain, a sequence of amino 
acids, known as a protein.
This protein folds into a complex, low-energy structure, and functions as
a cellular machine itself; the DNA sequence determines the amino acid sequence,
which in turn determines the folded structure of the protein.
This structure ultimately determines the protein's function within the cell.
Certain kinds of RNA also importantly function, sometimes with proteins, as cellular 
machines.
Thus, researchers are interested in DNA, RNA, and protein sequences, as well
as the structures that RNA and proteins form.
Researchers also study the resulting functions of RNA and proteins, in what
abundance they occur (this is known as gene expression), and how they interact
with each other (interaction networks).

Most but not all biological data take the form of \emph{sequence} data, either 
nucleotide sequences (nominally a four-letter alphabet representing the four 
DNA or RNA bases) or protein sequences (nominally a twenty-letter alphabet 
representing the twenty standard amino acids).
Sequence data are obtained in several ways, including mass spectrometry (which
can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence, 
and by inference the expression  of the gene it might translate).
However, the greatest volume of sequence data derive from next-generation 
sequencing (commonly abbreviated NGS) technologies, and comprise DNA sequences.
At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method.
Slow and labor intensive but well-tested, Sanger sequencing is still used to 
validate NGS results.
More recently, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple preparation
of samples.
Importantly to this Review, Illumina sequencing, as well as other approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing, does not read a single DNA
molecule end-to-end as one could a file from a hard drive.
Instead, DNA molecules are chopped into many small fragments;
from these we can measure 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads are read concurrently and must be reassembled in the correct order 
if an entire genome is to be read.
Current reads most typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
This so-called ``shotgun sequencing'' is analogous to the problem of 
reconstructing a large file from small, 
randomly-ordered reads, or a shredded paper document.
However, these NGS reads overlap, and they are many in number, so 
reconstructing an entire genome is feasible.
No sequencing technology is completely infallible, so sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many of these reads, typically stored in a FASTQ
file.

>>>>

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many biological 'omics` applications require us to store, access, 
and analyze large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have collaborated to bring the GATK 
(Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
Amazon Web Services are also commonly used for computational biology 
research and enterprise (e.g., DNANexus)~\cite{schatz2010cloud}.
However, while cloud computing frees researchers from maintaining their own
data centers, and provides cost-saving benefits when computing resources are
not needed continuously, it is no panacea.
First and foremost, cloud computing does not truly address the problem posed by
a corpus of data that is growing faster than Moore's law, because the computer
systems that make up those cloud datacenters are still themselves bound by
improvements in semiconductor technology; they still follow the trajectory
suggested by Moore's law.
Moreover, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.

@@

In the central dogma of molecular biology, DNA is transcribed into RNA, which
is translated by the ribosome into a polypeptide chain, a sequence of amino 
acids, known as a protein.
This protein folds into a complex, low-energy structure, which functions as
a cellular machine; the DNA sequence determines the amino acid sequence,
which in turn determines the folded structure of the protein.
This structure ultimately determines the protein's function within the cell.
Certain kinds of RNA also function, sometimes with proteins, as cellular 
machines.
Thus, researchers are interested in DNA, RNA, and protein sequences, as well
as the structures that RNA and proteins form.
Researchers also study the resulting functions of RNA and proteins, in what
abundance they occur (gene expression), and how they interact
with each other (interaction networks).

Most but not all biological data take the form of \emph{sequence} data, either 
nucleotide sequences (nominally a four-letter alphabet representing the four 
DNA or RNA bases) or protein sequences (nominally a twenty-letter alphabet 
representing the twenty standard amino acids).
Sequence data are obtained in several ways, including mass spectrometry (which
can determine protein sequence, as well as interactions) and RNA-seq (which can determine RNA sequence, 
and by inference the expression of the gene to which it might translate).
However, the greatest volume of sequence data derive from next-generation 
sequencing (NGS) technologies, and comprise DNA sequences.
At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method.
Slow and labor intensive but well-tested, Sanger sequencing is still used to 
validate NGS results.
More recently, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.
Importantly to this Review, Illumina sequencing, as well as other approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing, does not read a single DNA
molecule end-to-end as one could a file from a hard drive.
Instead, DNA molecules are chopped into many small fragments;
from these we can measure 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads are read concurrently and must be reassembled in the correct order 
if an entire genome is to be read.
Current reads most typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
This so-called ``shotgun sequencing'' is analogous to the problem of 
reconstructing a large file from small, 
randomly-ordered reads, or a shredded paper document.
However, these NGS reads overlap, and they are many in number, so 
reconstructing an entire genome is feasible.
No sequencing technology is completely infallible, so sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many of these reads, typically stored in a FASTQ
file.
